---
title: "Human progress and Energy"
author: "Declan Oberlies"
date: "2026-02-10"
output: html_document
---

```{r Load packeges, echo=FALSE, message=FALSE, warning=FALSE}
#loading packeges
library(readr)
library(dplyr)
library(ggplot2)
library(knitr)
library(plotly)
```

```{r energy, echo=FALSE, warning=FALSE, message=FALSE}
energy <- read_csv("../../data/IRENA data.csv", skip = 1)
#energy %>% head()
#View(energy)
```

```{r gapminder, echo=FALSE, message=FALSE, warning=FALSE}
library(gapminder)
#gapminder %>% head
#View(gapminder)
```

```{r carbon, echo=FALSE, warning= FALSE, message=FALSE}
url <- 'https://nyc3.digitaloceanspaces.com/owid-public/data/co2/owid-co2-data.csv'
  
carbon <-
  read_csv(url) 
#carbon %>% head
#View(carbon)
```

```{r econ, echo=FALSE, warning= FALSE, message=FALSE}
url <- 'https://raw.githubusercontent.com/ericmkeen/sewanee_esus/master/02_energy_sector/levelized-cost-of-energy.csv'
econ <- read_csv(url)
#econ %>% head()
#View(econ)
```

```{r eval=FALSE, include=FALSE}
econ%>% 
  filter(year == 2020) %>%
  group_by(source) %>%
  summarize(cost = mean(cost, na.rm = TRUE)) %>%
  kable

```
## Introduction

```{r fig.1, echo=FALSE, warning=FALSE, message=FALSE}
# Create summary data
fig_1_data <- gapminder %>%
  group_by(year) %>%
  summarize(
    mean_lifeExp = mean(lifeExp, na.rm = TRUE),
    mean_gdpPercap = mean(gdpPercap, na.rm = TRUE),
    .groups = "drop"
  )

# Calculate scale factor
scale_factor <- max(fig_1_data$mean_gdpPercap) / (max(fig_1_data$mean_lifeExp) + 10)

# Make graph
ggplot(fig_1_data, aes(x = year)) +
  geom_path(aes(y = mean_lifeExp), 
            color = "green", 
            size = 1) +
  geom_path(aes(y = mean_gdpPercap / scale_factor), 
            color = "orange", 
            size = 1) +
  scale_y_continuous(
    name = "Mean Life Expectancy (green)",
    sec.axis = sec_axis(
      ~ . * scale_factor,
      name = "Mean GDP Per Capita in USD (orange)"
    )
  ) +
  labs(
    title = "Human Progress",
    subtitle = "Advances in human standard of living since 1950",
    x = " "
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.title = element_text(size = 11)
  )
```

Since 1950, human standard of living has increased drastically. On average, people are living longer and earning more money. This is due to...

-   Advances in medicine and healthcare

-   Advances in childbirth procedures

-   Advanced manufacturing

...and much more. All this advancement and progress has come at a cost.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# carbon dataset
fig_2_data <- carbon %>%
  filter(country == "World") %>%
  group_by(year) %>%
  summarise(
    total_co2 = sum(cumulative_co2, na.rm = TRUE),
    .groups = "drop"
  )

#View(fig_2data)

## make graph
ggplot(fig_2data,
       aes(x = year, 
           y = totalCO2/1000)) +
  geom_point( size = .2) +
  geom_vline(xintercept = 1821, lty = 2.5 ) +
  annotate( geom = 'text',
            1816, 1000, angle = 90, label = 'first electric motor ') +
  geom_vline(xintercept = 1859, lty = 2.5 ) +
  annotate( geom = 'text',
            1854, 1000, angle = 90, label = 'first oil well ') +
  geom_vline(xintercept = 1882, lty = 2.5 ) +
  annotate( geom = 'text',
            1877, 1000, angle = 90, label = 'first power grid') +
  geom_vline(xintercept = 1950, lty = 2.5 ) +
  annotate( geom = 'text',
            1945, 1000, angle = 90, label = 'comericial plastic popularized') +
  labs(title = "Exponential Progress and Exponential Consiquences",
       y = "total CO2 emmited (thousands of tons)",
       x = " ") +
  theme_minimal()

```

The industry required to fuel our life stiles emit exponentially more CO~2~ every year. CO~2~ is a greenhouse gas, and thus an over abundance of it in our atmosphere can create positive feedback loops that threaten to destabilize global ecosystems, and even our agriculture. Many industries emit CO~2~, but the energy industry is by far the largest. As our energy use and carbon emissions increase, we need to be rethink how we produce and consume energy.

## Energy

Civilizations are defined by how they use energy, and human energy use is currently at an all time high. Three nations are particularly high in energy use, namely the USA, China, and India.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# energy dataset
#energy %>% pull(Year) %>% range
fig_3data <- data.frame(energy %>% 
  group_by(`Country/area`, Year) %>%
  summarize(total_energy = sum(as.numeric(`Electricity statistics`), na.rm = TRUE))) %>%
  filter(Year != 2024) %>%
  mutate(total_energy = total_energy/1000000) 

fig_3trace <- 
  fig_3data %>%
  filter(Country.area %in% c('United States of America (the)', 'China', 'India')) %>%
  rename(Country = Country.area) 

#gap_filt <- gapminder %>%
  #filter(year %in% c(2000:2023)) %>%
  #rename( Country.area = country, Year = year)
#View(gap_filt)
#fig_3data <- left_join(fig_3data, gap_filt, by = c("Country.area", "Year"))
#fig_3data %>%
  #mutate(energyPercap = (total_energy/pop)) 
# make a grpah
fig.3 <-
  ggplot(fig_3data,
       aes( x = Year,
            y = total_energy))+
  geom_point() +
  geom_path(data = fig_3trace, aes(color = Country), size =1) +
  labs(title = "Energetic Nations", y = "Total Energy (millions of GWh)")

ggplotly(fig.3)
```
*black dots represent countries, lines trace the progression of standout countries*

The USA became a manufacturing superpower in the first half of the 20th century. Since then it has outsourced much of its production, but still has heavy energy demands. 

China surpassed the USA in energy production in 2011. Due to its colossal population, china became a manufacturing superpower. As 
```{r echo=FALSE, message=FALSE, warning=FALSE}
# energy zoom in
#econ_filt <- 
  #econ %>%
  #filter( country %in% c('United States of America (the)', 'China', 'India'),
          #year %in% c(2000:2023))
#gap_filt <- gapminder %>%
  #filter(year %in% c(2000:2023), country %in% c('United States of America (the)', 'China', 'India'))
  
#fig_3trace <- 
  #fig_3trace %>%
  #rename(country = Country.area, year = Year)
#fig_4data <- left_join(fig_3trace, gap_filt, by = c("country", "year")) 
fig_4data <- energy %>%
  filter(`Country/area` %in% c('United States of America (the)', 'China', 'India')) %>%
  mutate(`Electricity statistics` = 
           as.numeric(na_if(`Electricity statistics`,"-"))) %>%
  mutate(`Electricity Generated` = `Electricity statistics`/1000) %>%
  filter(Technology != "Total renewable", Technology != "Total non-renewable")
  
#View(fig_4data)
# make a plot

fig.4 <- ggplot(fig_4data,
       aes(x = Year,
           y = `Electricity Generated`,
           color = Technology)) +
  geom_point() +
  geom_path() +
  facet_wrap(~`Country/area`, nrow = 3) +
  scale_y_continuous( breaks = c(1000, 3000,5000)) +
  labs( title = "",
        y = "Electricity Generation (thousands of GWh)") 

ggplotly(fig.4)
  
```


```{r eval=FALSE, warning=FALSE, include=FALSE}
ggplot(econ,
       aes(x = year, 
           y = cost,
           color = source)
       ) +
  geom_point() +
  geom_hline(yintercept = 0.05, lty = 2) +
  geom_hline(yintercept = 0.18, lty = 2) +
  annotate(geom= 'text',
           x=1990, 0.1, size = 5, label = 'Range of fossil\n fule costs')
  
```